### 2025-08-19 manuscript/manuscript.qmd
Revised text to MS

Academia suffers from issues regarding equity. A lack equity impacts productivity, innovation, and job satisfaction in the workplace, which hugely impact academic progress. One aspect of this inequity is that gender. Women are more at a disadvantage than men, with evidence suggesting that women are less present in positions of power. For instance, in the UK in 2016-17, 24.6% of professors are women [@bhopal2021competing] when compared to the 44% of all grades in the UK academia [@harris2025gender]. Similar situation holds for italy [@Filandri03082021], 34% in the US [@huang2020]. A a pay gap remains between men and women in academia (female-male wage ratio of 0.85 in 2020 in the UK) [@quadlin2023higher]. In a “publish or perish” culture of science and academia [@kiai2019protect], one crucial aspect to focus on is authorship. The number and quality (e.g., journal reputation) of authorships an individual determines crucial career development such as grant funding and future (permanent) positions.

Research across several scientific disciplines that show that women tend to be less represented in authors compared to men [@banks2025women]; @son2022scientific]. Encouragingly, however, there is a general trend that the ratio of women to men in authorships is improving. From around 1960 to 2021, studies show a general improvement [@gonzalez2020women; @ioannidis2023gender; @jemielniak2025gender]. However, there is evidence that this improvement is plateauing [@jemielniak2025gender] and that although women are more likely to be first author, women are still generally unfairly underrepresented as the last author [@gonzalez2020women; @rock2021quantifying; @shah2021gender]. There could be several reasons for such a disparity: despite efforts to give fair credit for author contributions [@ni2021gendered], for example, using the Contributor Roles Taxonomy (CRediT) system, assigning authorships can be unclear. What counts as an authorship may also differ between academic disciplines. Women are more likely to experience authorship disagreements [@ni2021gendered]. There are general disagreements of when to decide authorship - women preferring to discuss in earlier stages, but men choosing authorship at the final stage. Another reason could also come from parenthood and parental leave, which account for about ~40% of gender gap in career advancement [@nielsen2024getting]. Importantly, there are strides to improving this, with opinion papers discussing way to improve receipt and reporting of intellectual credit [@banks2025women] as well as creating equitable environments in academic science [@martinez2024comparison].

See @gonzalez2020women for a baseline from other disciplines:
- Hard Sciences (N=119,592) = 0.148
- Bio and Social Sciences (N=262,122) =  0.433
- Psychology (N=90,067) =  0.452

What is the situation in music psychology?
To improve certain academic equity issues, it is worthwhile focusing within disciplines and seeing how more local academic communities can provide more equity. To do this, it is worth giving a snapshot of specific fields. Indeed, instead of looking broadly at journal articles in STEM, or in several different kinds of journals, it can be worthwhile giving an idea of a more specific. For example, one paper explores authorship in Herpetology, showing that male authors outnumber female, but female authorship is increasing steadily across the last decade [@rock2021quantifying]. The current paper also focuses on music psychology as a discipline: within music psychology, there are many efforts to improve the field. There have been papers that provide a snapshot of the field to provide a rough overview of the state of affairs (e.g., @anglada2019visualizing) with suggestions of future directions to improve the field. [@eerola2023transparency] showcases how Open Science practices (e.g., preregistrations, sharing research materials, data, and analysis scripts) are relatively limited, and encourage authors in the field to start implementing such practices. @jakubowski2024 explores participant samples and musical stimuli used across a wide range of experiment to give an idea of how limited or generalizable the field is, which the view to see how – as a field – we need to diversify participant samples and stimuli to gain a more realistic understanding. In a similar way to these papers, the current paper aims to give a current overview of authorship in the field of music psychology.

@@

Our aim is to find out what is the gender distribution in the specialist journals of music psychology. What is the proportion of different types of authorships (lead, coauthors, and last authorships) for men and women in the published papers in the last 25 years? Are there specific trends in terms of countries of the affiliations, topics, or time?

>>>>

Academia suffers from issues regarding equity. A lack equity impacts productivity, innovation, and job satisfaction in the workplace, which hugely impact academic progress. One aspect of this inequity is that gender. Women are more at a disadvantage than men, with evidence suggesting that women are less present in positions of power. For instance, in the UK in 2016-17, 24.6% of professors are women [@bhopal2021competing] when compared to the 44% of all grades in the UK academia [@harris2025gender]. A similar situation holds for italy (24% female professors) [@Filandri03082021], and to some extent for the US (14%) [@spoon2023]. A pay gap between men and women in academia is also significant, (e.g., female-male wage ratio is 0.85 in 2020 in the UK) [@quadlin2023higher]. In a “publish or perish” culture of science and academia [@kiai2019protect], one crucial aspect to focus on is authorship. The number and quality (e.g., journal reputation) of authorships an individual determines crucial career development such as grant funding and future (permanent) positions.

Research across several scientific disciplines that show that women tend to be less represented in authors compared to men [@banks2025women; @son2022scientific]. Encouragingly, however, there is a general trend that the ratio of women to men in authorships is improving. From around 1960 to 2021, studies show a general improvement [@gonzalez2020women; @ioannidis2023gender; @jemielniak2025gender]. However, there is evidence that this improvement is plateauing [@jemielniak2025gender] and that although women are more likely to be first author, women are still generally unfairly underrepresented as the last author [@gonzalez2020women; @rock2021quantifying; @shah2021gender]. There could be several reasons for such a disparity: despite efforts to give fair credit for author contributions [@ni2021gendered], for example, using the Contributor Roles Taxonomy (CRediT) system, assigning authorships can be unclear. What counts as an authorship may also differ between academic disciplines. Women are more likely to experience authorship disagreements [@ni2021gendered]. There are general disagreements of when to decide authorship - women preferring to discuss in earlier stages, but men choosing authorship at the final stage. Another reason could also come from parenthood and parental leave, which account for about ~40% of gender gap in career advancement [@nielsen2024getting]. Importantly, there are strides to improving this, with opinion papers discussing way to improve receipt and reporting of intellectual credit [@banks2025women] as well as creating equitable environments in academic science [@martinez2024comparison].

It is important to recognise that fields of scholarship and academic disciplines vary considerably in terms of gender distribution [@huang2020]. For example, a large survey of authors by @gonzalez2020women showed that the hard sciences (N = 119,592) had the lowest prevalence of women (14.8%), whereas in the biological and social sciences (N = 262,122) the proportion was substantially higher (43.3%). In psychology, which is perhaps the closest benchmark for the present focus on music psychology, female authors accounted for 45.2% of the sample (N = 90,067).

What is the situation in music psychology?

To improve certain academic equity issues, it is worthwhile focusing on our discipline and to assess the equity in music psychology. The past efforts have already provided a snapshot a rough overview of the state of affairs (e.g., @anglada2019visualizing) with suggestions of future directions to improve the field. [@eerola2023transparency] showcases how Open Science practices (e.g., preregistrations, sharing research materials, data, and analysis scripts) are relatively limited, and encourage authors in the field to start implementing such practices. @jakubowski2024 explores participant samples and musical stimuli used across a wide range of experiment to give an idea of how limited or generalizable the field is, which the view to see how – as a field – we need to diversify participant samples and stimuli to gain a more realistic understanding. In a similar way to these papers, the current paper aims to give a current overview of authorship and gender patterns in the field of music psychology.

@@

Our aim is to find out what is the gender distribution in the specialist journals of music psychology. What is the proportion of different types of authorships (single, first, coauthors, and last authorships) for men and women in the published papers in the last 25 years? Are there specific trends in terms of countries of the affiliations, topics, or time?
